GRB Afterglows In The Multi-Messenger Era OR Electromagnetic Afterglows Of Structured Jets OR Structured Jets At All Angles

### 2019-05-06 structuredJets.tex
Updates

\title{GRB Afterglows In The Multi-Messenger Era}
\correspondingauthor{Geoffrey Ryan}
\email{gsryan@umd.edu}

@@

\begin{abstract}
	Gamma-ray bursts (GRBs) associated with gravitational wave events are, and will likely continue to be, viewed at a larger inclination than their fellow GRBs without gravitational wave detections.  As demonstrated by the afterglow of \gwbns{}, this requires an extension of the common GRB afterglow models which typically assume emission from an on-axis top hat jet.  We present a characterization of the afterglows arising from structured jets, utilizing both trans-relativistic numerical calculations and simple analytic approximations.  In many cases the temporal slope before the jet break is found to be a simple function of the ratio between the inclination angle and effective half opening angle of the jet.  Our models, both numerical and analytic, are publicly available as an open source Python package.

@@

GRB afterglows are commonly modeled as top hat blast waves with no internal structure pointed directly at the observer.  This has been largely sufficient given the data.  (Although: viewing angle effects can be measured (Ryan+ 2015)).  New observational capabilities have witnessed new phases of afterglow evolution, such as the discovery of plateaus with Swift.  Real jets (of course) have internal structure, but the top hat assumption has been a sufficiently spherical cow.

@@

%These errors quoted are numerical errors associated with the schemes used to solve Equations \ref{eq:Rdot}, \ref{eq:E}, and \ref{eq:thetadot} and perform the integral in Equation \ref{eq:flux2}.  That is, they are the errors relative to solving these equations perfectly.  The equations themselves are, however, an approximation themselves to the full physics of an afterglow light curve.  Comparison and calibration against analytic calculations and hydrodynamic simulations is necessary to ensure physically realistic results.  
%
%\begin{figure*}
%	\plotone{figs/boxfit_comp.pdf}
%	\caption{Top Panel: Comparison between top hat jet light curves from \afterglowpy{} (solid lines) and \boxfit{} (dashed lines). Bottom Panel: fraction difference between \afterglowpy{} and \boxfit{} light curves.  Four representative light curves are shown: radio on-axis ($\thobs=0$, $\nu=10^9$Hz, blue), radio off-axis ($\thobs=0.16$rad, $\nu=10^9$Hz, orange), x-ray on-axis ($\thobs=0$, $\nu=10^{18}$Hz, green), and x-ray off-axis ($\thobs=0.16$rad, $\nu=10^{18}$Hz, red).  Remaining parameters are shared: $\thC=0.1$rad, $\Eiso=10^{52}$erg, $\Eiso=10^{52}$erg, $n_0 = 10^{-3}$cm$^{-3}$, $p=2.2$, $\epse=10^{-1}$, $\epsB=10^{-2}$, $\dL=3.09 \times 10^{26}$cm, $z=0.028$. \label{fig:boxfitComp}}
%\end{figure*}
%
%Figure \ref{fig:boxfitComp} shows a comparison between top-hat light curves calculated with \afterglowpy{} and \boxfit{}.  The \boxfit{} code uses two-dimensional relativistic hydrodynamic simulations to fully capture the non-linear hydrodynamics of a decelerating blast wave and a ray tracing radiative transfer module to compute observed synchrotron light curves.  Figure \ref{fig:boxfitComp} shows the light curve of a $\thC=0.1$rad top-hat at radio and x-ray frequencies ($10^9$Hz and $10^{18}$Hz, respectively) with aligned ($\thobs = 0$) and misaligned ($\thobs=0.16$rad) viewing angles.  The overall agreement is good: \afterglowpy{} captures the salient features of both aligned and misaligned light curves including the jet break, spectral shape, and transition to the Sedov phase.  The relative error is less than 50\% before the onset of spreading.

@@

\begin{itemize}
	\item Comparison with other \gwbns{} fits
	\item Degeneracy between $E(\theta)$ and $\thobs$
	\item $g$ vs $\thobs / \thC$ plot with \gwbns{}
	\item Looking forward to O3
\end{itemize}

@@

Of these, \citet{Mooley:2018ab, Hotokezaka:2018aa, Ghirlanda:2019aa} include the apparent superluminal motion

@@

@@

@@

\section{The Jet Angular Profile}
\label{sec:jetAngularShape}

The angular structure of a jet, before deceleration, can be described by angular profiles of energy $\epsilon(\theta, \phi)$ and Lorentz factor $\Gamma(\theta, \phi)$. This structure is acquired by the jet at its initial formation at the central engine and then sculpted by its propagation through the merger ejecta.  Due to the complications of direct numerical simulation much theoretical work has focused on simple analytic prescriptions of jet structure, typically axisymmetric uniform top-hat, Gaussian, or power law (CITE Granot+02, G+K03, K+G03, Rossi+04).

We consider the latter two models, a Gaussian profile
\begin{equation}
	E = E_0 \exp\left( \frac{\theta^2 }{2 \theta_c^2}\right)\ , \label{eq:Gaussian}
\end{equation}
and a $\theta^{-2}$ power law
\begin{equation}
	E = \frac{E_0}{1 + \theta^2/\theta_c^2}\ . \label{eq:powerlaw}
\end{equation}
Both models are truncated at an angle $\theta_w$. It is assumed all observed emission occurs after the jet has entered the deceleration phase, so the Lorentz factor profile is simply $\Gamma \propto E^{1/2}$.

With recent progress in hydrodynamical codes and computer resources, numerical simulation of jet propagation from the central engine to the circumburst medium has become possible.  Figure \ref{fig:jetStructure} shows energy and Lorentz factor distributions from hydrodynamic simulations collected from the recent literature, with representative examples of the Gaussian and power law models. The Lorentz factor distributions show large variation, in part due to differing times of extraction from the numerical simulations.  We focus comparison on the energy profile, as this is largely fixed while the jet remains relativistic.

\begin{figure}
	\plotone{figs/jetStructure.pdf}
   	\figcaption{Energy and Lorentz factor profiles of jet models.  Profiles of numerical simulations are reported in Morsony+2007 (green), Mizuta+2009 (brown), Duffell+2015 (orange), Lazzati+2017 (pink), Kathirgamaraju+2017 (red), and Margutti+2018 (purple). The analytic boosted fireball model (Duffell+2015) is shown in blue.  Energy profiles for Morsony07, Mizuta09, and Duffell15 have been scaled to $dE/d\Omega(\theta=0) = 10^{52}$ erg/sr.  For comparison we also show representative power law ( Eq. \eqref{eq:powerlaw}, $\theta_c=1^\circ$, $\theta_w=60^\circ$, thick grey) and Gaussian (\eqref{eq:Gaussian}, $\theta_c=5.7^\circ$, $\theta_w=23^\circ$, thick black) profiles.  \label{fig:jetStructure}}
\end{figure}

There is agreement between the models that the jet energy is concentrated in a core of half-width $\lesssim 10^\circ$ with wings extending to larger angles.  Some models truncate at $\mathcal{O}(10)$ degrees, while others have large tails extending to near the equator.  The Gaussian and power law models respectively emulate this behavior, providing a simple parameterized representation of the landscape of numerical models.  Of particular note are the Lazzati 2017 and Margutti 2018 models, both of which were calculated specifically for \gwbns{} The core of the Margutti model is has a profile very near Gaussian, while the Lazzati model has a strong resemblance to a power law.

>>>>

\title{GRB Afterglows In The Multi-Messenger Era  OR  Electromagnetic Afterglows Of Structured Jets  OR  Structured Jets At All Angles}
\correspondingauthor{Geoffrey Ryan}
\email{gsryan@umd.edu}

@@

\author{Eleonora Troja}

\author{Luigi Piro}

\begin{abstract}
	Gamma-ray bursts (GRBs) associated with gravitational wave events are, and will likely continue to be, viewed at a larger inclination than their fellow GRBs without gravitational wave detections.  As demonstrated by the afterglow of \gwbns{}, this requires an extension of the common GRB afterglow models which typically assume emission from an on-axis top hat jet.  We present a characterization of the afterglows arising from structured jets, utilizing both trans-relativistic numerical calculations and simple analytic approximations.  In many cases the temporal slope before the jet break is found to be a simple function of the ratio between the viewing angle and effective half opening angle of the jet.  This provides a useful handle in the analysis of structured GRB afterglow light curves, but does not completely resolve the degeneracy between viewing angle and jet structure model.  Our models, both numerical and analytic, are publicly available as the open source \python{} package \afterglowpy{}.

@@

Gamma-ray burst (GRB) afterglows are commonly modeled as top hat blast waves with no internal structure pointed directly at the observer. This has been largely sufficient given the available data, although some evidence of viewing angle effects has been found \citep{Ryan:2015aa}.  New observational capabilities have witnessed new phases of afterglow evolution, such as the discovery of plateaus with Swift.  Real jets (of course) have internal structure, but the top hat assumption has been a sufficiently spherical cow.

@@

%These errors quoted are numerical errors associated with the schemes used to solve Equations \ref{eq:Rdot}, \ref{eq:E}, and \ref{eq:thetadot} and perform the integral in Equation \ref{eq:flux2}.  That is, they are the errors relative to solving these equations perfectly.  The equations themselves are, however, an approximation themselves to the full physics of an afterglow light curve.  Comparison and calibration against analytic calculations and hydrodynamic simulations is necessary to ensure physically realistic results.  
%
\begin{figure*}
	\plotone{figs/boxfit_comp.pdf}
	\caption{Top Panel: Comparison between top hat jet light curves from \afterglowpy{} (solid lines) and \boxfit{} (dashed lines). Bottom Panel: fraction difference between \afterglowpy{} and \boxfit{} light curves.  Four representative light curves are shown: radio on-axis ($\thobs=0$, $\nu=10^9$Hz, blue), radio off-axis ($\thobs=0.16$rad, $\nu=10^9$Hz, orange), x-ray on-axis ($\thobs=0$, $\nu=10^{18}$Hz, green), and x-ray off-axis ($\thobs=0.16$rad, $\nu=10^{18}$Hz, red).  Remaining parameters are shared: $\thC=0.1$rad, $\Eiso=10^{52}$erg, $\Eiso=10^{52}$erg, $n_0 = 10^{-3}$cm$^{-3}$, $p=2.2$, $\epse=10^{-1}$, $\epsB=10^{-2}$, $\dL=3.09 \times 10^{26}$cm, $z=0.028$. \label{fig:boxfitComp}}
\end{figure*}

Figure \ref{fig:boxfitComp} shows a comparison between top-hat light curves calculated with \afterglowpy{} and \boxfit{}.  The \boxfit{} code uses two-dimensional relativistic hydrodynamic simulations to fully capture the non-linear hydrodynamics of a decelerating blast wave and a ray tracing radiative transfer module to compute observed synchrotron light curves.  Figure \ref{fig:boxfitComp} shows the light curve of a $\thC=0.1$rad top-hat at radio and x-ray frequencies ($10^9$Hz and $10^{18}$Hz, respectively) with aligned ($\thobs = 0$) and misaligned ($\thobs=0.16$rad) viewing angles.  The overall agreement is good: \afterglowpy{} captures the salient features of both aligned and misaligned light curves including the jet break, spectral shape, and transition to the Sedov phase.  The relative error is less than 50\% before the onset of spreading.

TODO: FINISH THIS SECTION.

@@

@@

TODO: INCLUDE BRIEF COMPARISON TO EACH.  NOTE DIFFERENT MODELS PRODUCE SAME LIGHT CURVE, STRONG EFFECT OF VLBI.

@@

In the coming years there will be more \gwbns{}-like events, although perhaps few with as extensive followup campaigns.  Since many of these events may be faint and distant, it will be important to leverage as much information from the afterglow light curve as possible.  As shown here, and evidenced by the diversity of models that have ``fit'' the \gwbns{} afterglow, there is a large degeneracy between viewing angle and jet structure profile.  To make robust inferences about these events, several flexible jet models must be included in the analysis.

We hope \afterglowpy{} and similar products will aid these analyses in the future.

@@

We have constructed flexible models for the electromagnetic afterglows of angularly structured relativistic jets.  Through analytic approximations we can identify the basic phases of the structured afterglow light curve, determining formulae for the characteristic break time scales and closure relations.  We find the closure relations depend on a single parameter we call $g$ relating to the jet structure and viewing angle.  Measurements of $g$ itself are model independent, but relating it back to physical parameters of the jet is not.

We also constructed a numerical model of the afterglows of structured jets, implemented in the public \python{} package \afterglowpy{}. \afterglowpy{} computes afterglow light curves on the fly utilizing semi-analytic approximations to the jet evolution and synchrotron emission, taking into account relativistic beaming, the equal time of arrival surface, jet angular structure, trans-relativistic evolution, and jet spreading.  It is fast enough to be incorporated in MCMC parameter estimation routines.  Fitting the \gwbns{} afterglow with multiple \afterglowpy{} models reveals the degeneracy between jet structure and viewing angle.

@@
